---
title: "chl_nutrients"
output: html_document
date: "2023-12-12"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Nitrogen and Phosphorous are considered the main drivers of phytoplankton growth. *Characterize the relationship* Is there a statistically significant relationship between nitrogen, phosphourous, and chlorophyll-a over the 20 year period?

```{r}
# import libraries 
library(tidyverse)
library(here)
library(janitor)
library(naniar)
library(lubridate)
library(tsibble)
library(feasts)
library(zoo)
library(modelr)
```

## Load data

SBC LTER Ocean: Ocean Currents and Biogeochemistry: Nearshore water profiles (monthly CTD and chemistry), ongoing since 2000

```{r}
bottle <- read_delim(here("data", "LTER_monthly_bottledata_20220930.txt"), 
                     # BDL = below detection limit, so convert those to NA
                     na = "BDL") %>% 
  clean_names() %>% 
  # convert any 99999 value to NA values 
  replace_with_na_all(condition = ~.x == 99999) %>% 
  # two rows at station Q2 which is not in the meta data and a positive longitude...
  filter(station != "Q2")
```

## Can we see a long-term trend in chl-a?

```{r}
ggplot(data = bottle) +
  geom_point(aes(x = yyyy_mm_dd, y = chl_a_ug_l, color = target_depth_m)) +
  scale_color_viridis_c()
```

## What about with decomposition?

```{r}
# Average Chl-a measurements by month
bottle_chl_ts <- bottle %>% 
  mutate(yr_mo = yearmonth(yyyy_mm_dd)) %>% 
  group_by(yr_mo) %>% 
  summarize(mean_chl_a = mean(chl_a_ug_l, na.rm = TRUE)) %>% 
  # convert to tsibble for decomposition
  as_tsibble(index = yr_mo) %>% 
  # interpolate NA values so decomp works
  fill_gaps() %>% 
   mutate(mean_chl_a=if_else(is.nan(mean_chl_a), NA, mean_chl_a) %>% 
           na.approx())
  
# plot decomposition componenets
bottle_chl_ts %>% 
  model(STL(mean_chl_a)) %>% 
  components() %>% 
  autoplot() +
  labs(x = "Date",
       y = "Average Chlorophyll-a Concentration (\u03BCg/L)",
       title = "Chlorophyll-a Decomposition") +
  theme_bw()
```

Looks like there is a downward trend over time! Big bar, so small impact but

*But, seasonal component is definitely more important in driving overall variability.*

## More data wrangling to look at relationship

```{r}
bottle_nutrients <- bottle %>% 
  select(no2_no3_umol_l, ammonia_umol_l)

bottle_nutrients[is.na(bottle_nutrients)] <- 0

bottle_nutrients <- bottle_nutrients %>% 
  mutate(din_umol_l = no2_no3_umol_l + ammonia_umol_l)

bottle_nutrients$din_umol_l[bottle_nutrients$din_umol_l == 0] <- NA

bottle_complete <- bottle %>% 
  cbind(din_umol_l = bottle_nutrients$din_umol_l) %>% 
  mutate(n_p_ratio = din_umol_l/po4_umol_l) %>% 
  select(chl_a_ug_l, po4_umol_l, din_umol_l, n_p_ratio)
```

## OLS

```{r}
# multiple linear regression
summary(lm(chl_a_ug_l ~ po4_umol_l + din_umol_l, data = bottle_complete))
```

## Interaction term?

"Some of the effects of nitrogen are related to the effect on increasing the growth of plant tops and concurrently increasing the absorption of phosphorus. The ammonium form of nitrogen frequently increases phosphorus absorption more than the nitrate form." <https://www.sciencedirect.com/science/article/pii/S0065211308601273#>:\~:text=Some%20of%20the%20effects%20of,more%20than%20the%20nitrate%20form.

```{r}
# multiple linear regression with interaction term
summary(lm(chl_a_ug_l ~ po4_umol_l + din_umol_l + din_umol_l:po4_umol_l, data = bottle_complete))
```
### Interpretation

- intercept: the predicted concentration of chlorophyll-*a* when the concentration of both phosphate and dissolved inorganic nitrogen is 0 is 2.87 ug/L.
- po4_umol_l: The average change in chlorophyll-*a* concentration with an average one unit increase in phosphate concentration, when the dissolved inorganic nitrogen concentration is 0, is -0.84 ug/L.
- din_umol_l: The average change in chlorophyll-*a* concentration with an average one unit increase in dissolved inorganic nitrogen concentration, when the phosphate concentration is 0, is 0.16 ug/L.
- po4_umol_l:din_umol_l: the marginal effect of of phosphate and dissolved inorganic nitrogen on on each other's relationship with chlorophyll-*a*
  - A higher phosphate concentration increases the slope of the relationship between chlorohpyll-*a* and dissolved inorganic nitrogen, and vice versa.
  
#### Marginal effects
```{r}
devtools::source_url("https://raw.githubusercontent.com/jkarreth/JKmisc/master/ggintfun.R")
```

```{r}
mod <- lm(chl_a_ug_l ~ po4_umol_l + din_umol_l + din_umol_l:po4_umol_l, data = bottle_complete)
marginal_effects <- ggintfun(obj = mod, varnames = c("po4_umol_l", "din_umol_l"), 
         varlabs = c("Phosphate", "Dissolved Inorganic Nitrogen"),
         title = FALSE, rug = TRUE,
         twoways = TRUE)
ggsave("marginal_effects.png", plot = marginal_effects, width = 10, height = 6)
```


## Collinearity

```{r}
ggplot(data = bottle_complete, aes(x = din_umol_l, y = po4_umol_l)) +
  geom_point(alpha = 0.5) +
  labs(x = "Dissolved Inorganic Nitrogen (\u03BCmol/L)",
       y = "Phosphate (\u03BCmol/L)") +
  geom_smooth(method = lm, se = FALSE,
              color = "darkseagreen") +
  scale_x_continuous(expand = c(0, 0), limits = c(0, 25)) +
  scale_y_continuous(expand = c(0,0), limits = c(0, 3)) +
  theme_bw()
```


## further work
- add seasonality
- site and depth

# Outliers
looking at the chl-a distribution, there are some definitte outliers that are skewing the distribution
```{r}
hist(bottle_complete$chl_a_ug_l, breaks = 20)
```
What about the distribution of the nutrients?

```{r}
hist(bottle_complete$din_umol_l)
hist(bottle_complete$po4_umol_l)
```
Since they all have a positive skew, let's try taking the log of them.

```{r}
bottle_logs <- bottle_complete %>% 
  mutate(log_chla = log(chl_a_ug_l),
         log_din = log(din_umol_l),
         log_po4 = log(po4_umol_l))
```
Let's look at the distribution of the logs
```{r}
hist(bottle_logs$log_chla)
hist(bottle_logs$log_din)
hist(bottle_logs$log_po4)
ggplot(data = bottle_logs, aes(x = log_din, y = log_po4)) +
  geom_point(alpha = 0.5)
```
## q-q plots
```{r}
# chl
ggplot(bottle_logs, aes(sample = log_chla)) +
  geom_qq(color = 'darkgreen') +
  geom_qq_line() +
  labs(title = "chl Q-Q Plot") +
  theme_bw()

# po4
ggplot(bottle_logs, aes(sample = log_po4)) +
  geom_qq(color = 'darkblue') +
  geom_qq_line() +
  labs(title = "po4 Q-Q Plot") +
  theme_bw()

# din
ggplot(bottle_logs, aes(sample = log_din)) +
  geom_qq(color = 'purple') +
  geom_qq_line() +
  labs(title = "din Q-Q Plot") +
  theme_bw()
```


## mulitple linear regression w/ interaction term with logs
```{r}
# multiple linear regression with interaction term
summary(lm(log_chla ~ log_po4 + log_din + log_din:log_po4, data = bottle_logs))
```
### marginal effects
```{r}
model_log <- lm(log_chla ~ log_po4 + log_din + log_din:log_po4, data = bottle_logs)
ggintfun(obj = model_log, varnames = c("log_po4", "log_din"), 
         varlabs = c("Log Phosphate", "Log Dissolved Inorganic Nitrogen"),
         title = FALSE, rug = TRUE,
         twoways = TRUE)
```


```{r}
ninety_nine_pctile <- quantile(bottle_complete$chl_a_ug_l, .99, na.rm = TRUE)
bottle_test <- bottle_complete %>% 
  mutate(chl_a_ug_l = ifelse(chl_a_ug_l > ninety_nine_pctile, ninety_nine_pctile, chl_a_ug_l))

ggplot(data = bottle_test) +
  geom_point(aes(x = din_umol_l, y = chl_a_ug_l))

ggplot(data = bottle_test) +
  geom_point(aes(x = po4_umol_l, y = chl_a_ug_l))
```
```{r}
summary(lm(chl_a_ug_l ~ po4_umol_l + din_umol_l + din_umol_l:po4_umol_l, data = bottle_test))
```
```{r}
plot(lm(chl_a_ug_l ~ po4_umol_l + din_umol_l + din_umol_l:po4_umol_l, data = bottle_complete))
```

